Fail Faster

### 2025-03-18 paper/paper.tex
progress

\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}

\usepackage{listings}
\usepackage{subcaption}
\usepackage{xspace}
\lstset{
  mathescape=true,
  frame=none,
  xleftmargin=10pt,
  stepnumber=1,
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  % language=haskell,
  keepspaces=true,
  tabsize=2,
  emphstyle={\bf},
  % commentstyle=\it\color{dkgreen},
  stringstyle=\mdseries\ttfamily,
  showspaces=false,
  keywordstyle=\bfseries\ttfamily,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  showstringspaces=false,
  % morecomment=[l]\%,
  % moredelim=**[is][\color{dkgreen}]{@}{@}
}

@@

Property-based testing (PBT) is a technique for software testing which uses
random test inputs to evaluate logical properties that the system is expected to
satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This workflow makes
sense because the developer who just wrote the code is often the best person to
fix a potental bug; if a bug is found later, it will need to be fixed by someone
with less context. \jwc{Not sure how the previous sentence is relevant. It delays us getting to the point.}  All of this means that PBT is often run with very short time
budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
that bugs can be found in that time-frame.
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}

Given these strict timelines, it is important to think of PBT as
performance-sensitive process. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning. \jwc{... to find bugs fast}

We find that the performance of existing generator libraries falls short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder the compiler's optimizations, shifting more work to
run time.
Second, the \rand that underlies the generator's operation --- the source of
random numbers that are used to make choices in the generator --- 
can constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
These observations together raise a fundamental question: can generator
libraries produce code that is both expressive \textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To address this question, we propose a two-pronged approach for designing
generator libraries that preserves their expressiveness while eliminating
unnecessary overhead.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing idiomatic functional DSLs~\cn.
We extend this technique to generators \jwc{``apply this technology to''? We're not extending staging, we're a client}, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \hg{More words here for
parallelism?} Together, these insights provide a recipe for improving even
state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, a
high-quality PBT library in OCaml, authored by Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons, \propose{as} differences in \proposechange{generation speed}{bugfinding power} are
\proposechange{directly}{solely} attributable to the staging and \rand optimizations.

We evaluate the optimizations in \name with a series of case studies in each
generator library, and show that \name runs faster than \bq.
Further, we show that this performance benefit\hg{wc?} translates to bug-finding
ability by running these case studies in the Etna framework~\cite{etna}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.
Finally, to demonstrate the generality of our technique, we implement a version
of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators
in the Scala port of the library---thereby highlighting its portability to other
strict functional languages.
\iflater
\jwc{I think we should more explicitly point out the portability earlier, maybe
up where we say we're applying it to BQ.}\hg{Agree}
\fi

In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead \jwc{Swap}. \secref{section:motiv}
    \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
    \item We implement a version of \bq, \name, that showcases our optimization technique \jwc{Calling it an ``optimization technique'' begs the ``so they just made a thing faster'' question} while maintaining strict program equality. \secref{section:4}
    \jwc{We implement \emph{two} versions of \name{}! Once for BQ and once for ScalaCheck.}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}

@@

\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
To establish the equivalence of generators up to program equality, we use a PBT harness to compare their outputs. 
Specifically, we generate random sizes and seeds and evaluate whether the generators under test produce equal results. 
If the equality property fails, we report the unequal trees; if, over 1,000 trials, no ``counterexample'' is reported, we consider the
behavior of the generators to be equivalent.
\subsection{Benchmarking speed \& resource usage}

\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints without bind (use map): lots of sampling, minimal binds.
    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
  \end{itemize}
}
\subsection{Impact on bug-finding ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques. 
Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value that violates that ordering.

@@

We evaluate each case study with both type-derived and custom generators.
Each generator has four implementations, which are equivalent up to program equality:
\begin {itemize}
  \item The original \bq generator (``BQ'').
  \item A staged version of the generator (``BQ++'').
  \item A staged version of the generator using the C \rand (``BQC++'').
  \item A staged version incorporating the CSR \rand (``BQCSR++'').
\end{itemize}

@@

\subsubsection{Variance between seeds}
The time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this is true in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup.
To account for these edge cases, we repeated the above process for 10 seeds and computed the geometric average of their speedups.

@@

>>>>

\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.17}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{xspace}
\lstset{
  mathescape=true,
  frame=none,
  xleftmargin=10pt,
  stepnumber=1,
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  % language=haskell,
  keepspaces=true,
  tabsize=2,
  emphstyle={\bf},
  % commentstyle=\it\color{dkgreen},
  stringstyle=\mdseries\ttfamily,
  showspaces=false,
  keywordstyle=\bfseries\ttfamily,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  showstringspaces=false,
  % morecomment=[l]\%,
  % moredelim=**[is][\color{dkgreen}]{@}{@}
}

@@

Property-based testing (PBT) is a software testing technique which uses
random test inputs to validate logical specifications.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners often interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This means that PBT is often run with very short time
budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
that bugs can be found in that time-frame.\iflater
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}\fi

Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast. \iflater\bcp{Incorporate fail fast somehow.}\fi

However, careful measurements show that the performance of existing generator libraries falls significantly short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder compiler optimizations, shifting work to
run time.
Second, the \rand that underlies the generator's operation---the source of the
random numbers that are used to make choices in the generator---can 
constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To address this question, we propose a two-pronged approach for designing
generator libraries that preserves their expressiveness while eliminating
unnecessary overhead.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

@@

To demonstrate the effectiveness of our approach, we apply it to \bq, a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{100\% semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The semantic equivalence between \name and \bq allows us to 
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

We evaluate the optimizations in \name with a series of case studies in each
generator library, and show that \name runs faster than \bq.
Further, we show that this performance benefit\hg{wc?} translates to bug-finding
ability by running these case studies in the Etna framework~\cite{etna}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.
Finally, to demonstrate the generality of our technique, we implement a version
of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators---thereby highlighting its portability to other
strict functional languages.
\iflater
\jwc{I think we should more explicitly point out the portability earlier, maybe
up where we say we're applying it to BQ.}\hg{Agree}
\fi

In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, and abstraction overhead, and \rand choice. \secref{section:motiv}
    % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
    \item We implement our approach in OCaml and Scala, showcasing its cross-language applicability.
    \secref{section:4}
    \item We construct generators in \name and \scalaname, and show that they give substantial performance improvements. Finally, we show
    that these performance improvements extend to bug-finding ability.
     \secref{section:eval}
\end{enumerate}

@@

\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
To establish the equivalence of generators up to program equality, 
we employ a PBT harness to compare their outputs over 1,000 random seeds. 
If a discrepancy arises, we report the unequal trees; if no counterexample is found,
we conclude that the generators exhibit equivalent behavior.
\subsection{Benchmarking speed \& resource usage}

\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints without bind (use map): lots of sampling, minimal binds.
    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
  \end{itemize}
}
\subsection{Impact on bug-finding ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques. 
Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.

@@

We evaluate our case studies with both type-derived and custom generators.
Each generator has four implementations, which are equivalent up to program equality:
\begin {itemize}
  \item The original \bq generator (``BQ'').
  \item A staged version of the generator (``Staged'').
  \item A staged version of the generator using the C \rand (``Staged C'').
  \item A staged version incorporating the CSR \rand (``Staged CSR'').
\end{itemize}

@@

\subsubsection{Variance between seeds}
The time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.

@@

\begin{figure}[h]
  \centering
  \vspace{0.5em} % Kept the vertical spacing
  \begin{tabular}{lrl}
  \toprule
  Benchmark & \multicolumn{2}{c}{avg. time (ns/op) $\pm$ std. dev.} \\
  \midrule
  BST & $3,560.813$ & $\pm\>1,511.201$ \\
  BST, Staged & $751.287$ & $\pm\>39.890$ \\
  BST (Type-Derived) & $44,619.863$ & $\pm\>5,761.568$ \\
  BST (Type-Derived), Staged & $6,863.209$ & $\pm\>416.188$ \\
  RBT (Type-Derived) & $60,559.846$ & $\pm\>4,901.980$ \\
  RBT (Type-Derived), Staged & $8,511.528$ & $\pm\>324.615$ \\
  STLC & $44,497.623$ & $\pm\>1,970.484$ \\
  STLC, Staged & $7,151.259$ & $\pm\>355.639$ \\
  \bottomrule
  \end{tabular}
 \caption{\todo{Understand Joe's implementation of this...}}
 \label{fig:benchmark-table}
 \end{figure}
